---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ScatterGeo

Use this data source: https://ndownloader.figshare.com/files/5360966

Find out about the libraries here: http://www.htmlwidgets.org/showcase_leaflet.html

## Leaflet

This code chunk shows how to import the data from figshare.
```{r, echo=TRUE}
University_Data <- read.csv(file= "https://ndownloader.figshare.com/files/5360966")
colnames(University_Data) <- c("University", "Latitude", "Longitude")
```


This code chunk shows how I first got rid of row 14 as it was a duplicate of another row.
```{r, echo=TRUE}
University_Data2 <- University_Data[1:13,]
University_Data2 
End_Uni <- University_Data[15:16,]
university_data <- rbind(University_Data2, End_Uni)
```

This code chunk shows how I added a column to show the URL and Guardian league standing of each university.
```{r, echo=TRUE}
##Adding the URL column
URL<- c("http://www.ox.ac.uk/", "https://www.dur.ac.uk/", "http://www.qub.ac.uk/", "https://www.leeds.ac.uk/", "http://www.uwe.ac.uk/", "http://www.kcl.ac.uk/", "http://www.ed.ac.uk/", "http://www.ucl.ac.uk/", "http://www.sheffield.ac.uk/", "http://www.gla.ac.uk/", "https://www.kent.ac.uk/",  "https://www.cam.ac.uk/", "http://www2.gre.ac.uk/", "http://www.birmingham.ac.uk/", "http://polytech.univ-tours.fr/" )
university_data <- cbind(university_data, URL)

##Adding the league standings column
league_standings <- c(2, 6, 48, 16, "N/A", 42, 22, 14, 41, 26, 23, 1, 98, 13, "N/A")
university_data <- cbind(university_data, league_standings)

```

This code chunk shows the function I made for my popups to be used later.
```{r, echo=TRUE}
popup_func <- function(a,b,c) {
paste0(
"<b><a href=",b,">",a,"</a></b>",
"<p>Guardian League Standing #",c,"</p>"
)}
```

This code chunk shows how I used the library leaflet to create an interactive graphs including popups using my function from earlier.
```{r, echo=TRUE}
library(leaflet)
pal <- colorQuantile("YlOrRd", NULL, n=8)
leaflet(university_data) %>% addTiles() %>% addCircleMarkers(popup = ~popup_func(University,URL,league_standings))
```

##plotly

```{r}
library(plotly)

m <- list(
  colorbar = list(title = "1"),
  size = 8, opacity = 0.8, symbol = 'square'
)


g <- list(
  scope = 'UK',
  projection=list(type='UK'),
  showland = TRUE
 
)
plot_ly(university_data,
        type = 'scattergeo', 
        lat=Latitude, 
        lon=Longitude, 
        mode="markers") %>%
  layout(title = 'Uk universities<br>(hover for university)')



plot_ly(university_data, lat=latitude, lon=longitude, text=hover, color=cnt, , locationmode= 'UK', mode=markers, marker=m)%/%
  layout(title = 'Uk universities<br>(hover for university)', geo=g)

```

