---
title: "Interactive Tables"
author: "Martin John Hadley"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(htmltools)
```

# Interactive Tables Tutorial

This tutorial demonstrates how to create interactive tables like the one below using the [DT](https://rstudio.github.io/DT/) and [rhandsontable](http://jrowen.github.io/rhandsontable/#introduction) libraries, like the one below (built using `DT`). Note that there are three interactive shiny apps associated with this tutorial:

- [Comparison of `DT` and `rhandsontable`](https://livedataoxford.shinyapps.io/htmlwidget_template_datatables)
- Server-side 

```{r, echo = FALSE}
library(DT)
library(lubridate)
desktopItems <- read.csv(file = "https://ndownloader.figshare.com/files/5360960")
desktopItems$Timestamp <- mdy_hms(desktopItems$Timestamp, truncated = 2)
datatable(desktopItems,
          filter = 'top', rownames = FALSE,
          options = list(pageLength = 5, autoWidth = TRUE)) %>% formatDate( ~ Timestamp, method = "toUTCString")
```

These libraries are useful for visualising almost all rectangular datasets, with support for a variety of data types (dates, strings, URLS, etc).

## Import and Display Data

The data for this template is a .csv file accessed from Figshare [here](https://figshare.com/articles/Investigation_into_the_number_of_desktop_items_on_University_staff_computers_broken_down_by_department_and_operating_system_/3425729) using `read.csv`. 

```{r import_data}
desktopItems <- read.csv(file = "https://ndownloader.figshare.com/files/5360960", stringsAsFactors = F)
knitr::kable(head(desktopItems))
```

The first columns contain dates which need to be converted into a date format, for which we use the `lubridate` library. Unfortunately, the dates are slightly complicated - they are formatted `mm/dd/yyyy hh:mm:ss` but not all entries have seconds. The argument `truncated=2` allows us to handle these dates easily:

```{r}
library(lubridate)
desktopItems$Timestamp <- mdy_hms(desktopItems$Timestamp, truncated = 2)
knitr::kable(head(desktopItems))
```

# DT

The `DT` library binds to the extremely popular JavaScript library, <a href="datatables.net">datatables.net</a>. It is pretty much the defacto standard for interactive (responsive) tables, they have the added advantage of being able to paginate data in a variety of ways (including on a server-side engine). Any data.frame can be provided to datatables, and little effort must be put into pre-processing data ready for display in the table.

This data.frame can now be provided to `datatable` from the `DT` library for display, but note the poor formatting of the timestamp

```{r}
library(DT)
datatable(desktopItems)
```

## Column Formatting

There are a number of utility functions for formatting columns:

- formatCurrency
- formatPercentage
- formatRound
- formatDate

In the case of date, there are a number of different formats available, which are hidden in the symbol `DateMethods`, we will use `toUTCString` to format the data nicely.

```{r}
DT:::DateMethods
```

The `format*()` functions must be provided with both a datatable to operate on and a which columns to format, it is therefore convenient to use the `%>%` operator:

```{r}
datatable(desktopItems) %>% formatDate(~Timestamp, method = "toUTCString")
```


## Column Filters

The default datatable output includes a box in the top-right corner labelled "search" that allows the entire table to be filtered, but it's often convenient to enable per column filtering. This is particularly useful for dates, the "Timestamp" column can now be filtered according to a date range. In addition to the column filters, the argument `pageLength` has been added to `options` to reduce the scrolling required to navigate through this tutorial.

```{r}
datatable(desktopItems,
          filter = 'top',
          options = list(pageLength = 5)) %>% formatDate( ~ Timestamp, method = "toUTCString")
```

Please note that at present there is not a simple, non-JavaScript method known for formatting the date range filter.

### Responsive Reflow

```{r}
datatable(desktopItems,
          filter = 'top',
          options = list(pageLength = 5, responsive = TRUE, autoWidth = TRUE)) %>% formatDate( ~ Timestamp, method = "toUTCString")
```

## Server Side Processing

Often datatables are used to present a small amount of data from a large (probably multiple tabled) database, in such cases it's important not to download the entire dataset into the client's browser and instead to rely on the server to paginate the data. The following example is from <a href="http://rstudio.github.io/DT/server.html">http://rstudio.github.io/DT/server.html</a>.

There is a JSONP dataset provided by datatables.net that is setup to paginate server side here: https://datatables.net/examples/server_side/scripts/jsonp.php. In order to populate a datatable with the data we must build a `data.frame` with the right structure to contain the data:

```{r}
container_df <- data.frame(
  "First name" = character(),
  "Last name" = character(),
  "Position" = character(),
  "Office" = character(),
  "Start date" = character(),
  "Salary" = numeric(),
  check.names = FALSE
  )
```

Datatable is instructed to use ajax to pull the data server side from the JSONP data source as follows:

```{r}
datatable(container_df, rownames = FALSE, options = list(
  pageLength = 5,
      ajax = list(
        serverSide = TRUE, processing = TRUE,
        url = 'https://datatables.net/examples/server_side/scripts/jsonp.php',
        dataType = 'jsonp'
      )
    ))
```

# rhandsontables

The `rhandsontables` library binds to the [handsontable](https://handsontable.com/) which is designed for displaying interactive spreadsheets - with a focus on editing. Note that the handsontable.JS license is licensed such that educational use is free, but commercial use is not - refer to https://handsontable.com/ for details.

The data.frame storing our data can be provided to `rhandsontable` for display, note that values in the table can be updated freely.

```{r}
library(rhandsontable)
rhandsontable(desktopItems)
```

Editing can be turned off globally or for individual columns with `hot_col` and the `readOnly` argument, as with most `htmlwidgets` the `%>%` ("pipe") operator is used to modify existing output:

```{r}
rhandsontable(desktopItems) %>% hot_col("Desktop.Items", readOnly = TRUE)
```

In this document we are only concerend with displaying data, we therefore do not discuss further customisation of `rhandsontable` beyond highlighting the selected row and column with the `highlight*` argument to `hot_table` and enabling sorting of columns (double-click on headings)

```{r}
rhandsontable(desktopItems) %>% hot_col("Desktop.Items", readOnly = TRUE) %>% 
  hot_table(highlightCol = TRUE, highlightRow = TRUE) %>% 
  hot_cols(columnSorting = TRUE)
```

## Shiny App

A shiny app containing interactive versions of the datatables above is available here: <a href="https://livedataoxford.shinyapps.io/htmlwidget_template_datatables/">https://livedataoxford.shinyapps.io/htmlwidget_template_datatables/</a>.

The following types of interaction are supported:

- Switch between visualisation library
- Select rows in the interactive table and see the row number displayed in the app



