---
title: "htmlwidgets Timelines"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This tutorial demonstrates how to create interactive timelines (or Gantt charts) like the one below using a variety of different libraries, currently including; ggplot2, plotly and googleVis.

```{r, echo=FALSE, fig.height=1, message=FALSE, warning=FALSE}
library(googleVis)
library(shiny)
shinyApp(
  ui = fluidPage(
    htmlOutput("gvis_timeline"),
    "foo"
  ),
  server = function(input, output){
    datTL <- data.frame(Position=c(rep("President", 3), rep("Vice", 3),"President"),
                    Name=c("Washington", "Adams", "Jefferson",
                           "Adams", "Jefferson", "Burr","Washington"),
                    start=as.Date(x=c(rep(c("1789-03-29", "1797-02-03", 
                                          "1801-02-03"),2),"1801-02-03")),
                    end=as.Date(x=c(rep(c("1797-02-03", "1801-02-03", 
                                        "1809-02-03"),2),"1809-02-03")))
    
    output$gvis_timeline <- renderGvis({dat <- data.frame(Room=c("Room 1","Room 2","Room 3"),
                  Language=c("English", "German", "French"),
                  start=as.POSIXct(c("2014-03-14 14:00", "2014-03-14 15:00",
                                     "2014-03-14 14:30")),
                 end=as.POSIXct(c("2014-03-14 15:00", "2014-03-14 16:00",
                                  "2014-03-14 15:30")))
gvisTimeline(data=dat, rowlabel="Language",
                   start="start", end="end")})
  }, options = list(height = "200px")
)
```

The datasets that this tutorial considers are structured as follows:

```{r example_df, echo=FALSE}
example_df <- data.frame(
  "Start Date" = c("1789-03-29", "1797-02-03", "1801-02-03"),
  "End Date" = c("1797-02-03", "1801-02-03", "1809-02-03"),
  "Timeline Label" = c("Thing A","Thing B","Thing A"),
  "Timeline Tooltip Info" = c("Label 1", "Label 2", "Label 2"),
  "Event Category" = c("X","X","Y")
)
DT::datatable(example_df)
```

Where the "Start Date" and "End Date" column contain the start and end dates for the events, which must be formatted as YYYY-MM-DD. If your dates are not formatted like this, then refer to the "Date Manipulations Script" provided separartely. The "Timeline Label" and "Timeline Tooltip Info" columns provide information about how the events should be labelled in the timeline and what should be shown in the tooltip, respectively. Finally, the "Event Category" column is used for colour coding events.

Note that this template covers both how to build gantt charts inside of an HTML RMarkdown file and how to functionalise the code so as to conveniently switch between different categories and metrics in a Shiny app.

## Import Data

Please use this dataset - https://ndownloader.figshare.com/files/5409008 - which is described here - https://figshare.com/articles/Collated_Datasets_for_Oxford_University_Live_Data_Project/3444278

Owen: Make an offhanded joke about us using the "Political Party" column as both the "Timeline Tooltip Info" and "Event Category" data columns so as to not make this tutorial party political. ... Or if you can think of adding a column to the dataset based on what is already available then please do so.

## ggplot

`ggplot` is a great library for creating charts.

```{r}
## This imports a sample data set I provided to you
imported_timeline_data <- read.csv(file = "timeline_data.csv")
timeline_data <- imported_timeline_data
timeline_data$Valid.From <- as.Date(timeline_data$Valid.From)
timeline_data$Valid.To <- as.Date(timeline_data$Valid.To)
library(ggplot2)

ggplot(timeline_data, aes(x=Valid.From, xend=Valid.To, y=Name.of.Policy, yend=Name.of.Policy, colour=Type)) +
         geom_segment(size=3) + 
  xlab("Date") + ylab("Name of Policy")
```

## plotly

Plotly lets you convert ggplot plots directly into interactive charts.

```{r}
library(plotly)
ggplotly(ggplot(timeline_data, aes(x=Valid.From, xend=Valid.To, y=Name.of.Policy, yend=Name.of.Policy, colour=Type)) +
         geom_segment(size=3) + 
  xlab("Date") + ylab("Name of Policy"))
```


## gvisTimeline

Note that some charts from the `googleVis` library depend on server-side operations from Google, and therefore cannot be displayed within a standard RMarkdown HTML page - `gvisTimeline` is a good example of this. In order to display the timeline is is necessary to place the timeline in either an iframe or within a Shiny app. This tutorial makes the choice of displaying within a Shiny application as iframes would require you to look into hosting your output `gvisTimeline` elsewhere, and that's needlessly complicated.

```{r, echo=TRUE}
library(googleVis)
library(shiny)
shinyApp(
  ui = fluidPage(
    htmlOutput("gvis_timeline")
  ),
  server = function(input, output){
    datTL <- data.frame(Position=c(rep("President", 3), rep("Vice", 3),"President"),
                    Name=c("Washington", "Adams", "Jefferson",
                           "Adams", "Jefferson", "Burr","Washington"),
                    start=as.Date(x=c(rep(c("1789-03-29", "1797-02-03", 
                                          "1801-02-03"),2),"1801-02-03")),
                    end=as.Date(x=c(rep(c("1797-02-03", "1801-02-03", 
                                        "1809-02-03"),2),"1809-02-03")))
    
    output$gvis_timeline <- renderGvis({
      gvisTimeline(data=datTL, rowlabel="Name",
                         barlabel="Position",
                         start="start", 
                         end="end",
                         options=list(timeline="{groupByRowLabel:true}",
                                      backgroundColor='#ffd'))
    })
  }, options = list(height = "200px")
)
```

```{r}
timeline_data2 <- read.csv(file="https://ndownloader.figshare.com/files/5409008")
timeline_data2
timeline_data2$Start.Date <- as.Date(timeline_data2$Start.Date)
timeline_data2$End.Date <- as.Date(timeline_data2$End.Date)
```


##Prime Minister Timeline
```{r}
ggplot(timeline_data2, aes(x=Start.Date, xend=End.Date, y=Prime.Minister[1:24], yend=Prime.Minister[24], colour=Political.Party)) + geom_segment(size=1) + 
  xlab("Date") + ylab("Prime Minister")
```

```{r}
prime_minister_data <- prime_minister_data[,1:4]
prime_minister_data
```


```{r}
library(googleVis)
library(shiny)
shinyApp(
  ui = fluidPage( htmlOutput("gvis_timeline")
                  ),
  server = function(input,output) { 
output$gvis_timeline <- renderGvis({
  gvisTimeline(data=prime_minister_data, rowlabel = "Prime Minister")
  barlabel="Position"
  start="start"
end="end"
options=list(timeline="{groupByRowLabel:true}",
             backgroundColor='#ffd')
})}, 
options=list(height="200px")
)