---
title: "htmlwidgets"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Scope of Document

This document covers those htmlwidgets that are also on CRAN. The full list of these is available here: https://cran.r-project.org/web/packages/htmlwidgets/index.html

Currently covers: 

```{r}
htmlwidgets_on_CRAN <- c("d3heatmap", "DiagrammeR", "diffr", "DT", "dygraphs", "edgebundleR", 'formattable', 'ggiraph', 'gSEM', 'highcharter', 'leaflet', 'mapview', 'metricsgraphics', 'networkD3', 'pairsD3', 'plotly', 'qrage', 'qtlcharts', 'radarchart', 'rAmCharts', 'rbokeh', 'rchess', 'rglwidget', 'rhandsontable', 'rpivotTable', 'scatterD3', 'svgPanZoom', 'threejs', 'tmap', 'treescape', 'trelliscope', 'vegalite', 'visNetwork')
```


## Packages not yet on CRAN

These are pretty exciting but not on CRAN yet 

https://github.com/ThomasSiegmund/D3TableFilter



# What is htmlwidgets

A way to easily create bindings R bindings to existing JavaScript libraries, this document primarily considers the most established libraries utilising the htmlwidgets frameworks.

## Creating Your Own

It is fairly easy, in fact here's a guy that dedicated to build 1 htmlwidgets library a week for a whole year: http://www.buildingwidgets.com/blog/2015/1/2/can-i-commit

The following example allows us to create a GIF in the browser - which could be embedded into a Shiny app.

```{r}
devtools::install_github("timelyportfolio/gifrecordeR")
library(gifrecordeR)

library(htmltools)

browsable(
  tagList(list(
    tags$div(
      id = "div_record",
      style = "width: 400px; height: 400px;",
      tags$h2(
        contentEditable = "true",
        "Recording Live ... (I'm editable)"
      ),
      tags$p(
        contentEditable = "true",
        "Feel free to change me.  Careful though it might be
        recorded in a GIF."
      )
    ),
    gifrecord("#div_record")
  ))
)

```

# High Level Library Comparisons

Basic comparison of htmlwidgets libraries, 

## Chart Types

Generic Chart types:

note that "diagrams" means flowcharts and generic diagrams, for instance this periodic table created using `rbokeh` from http://hafen.github.io/rbokeh
```{r rbokeh_periodic_table}
## Copied verbatim from http://hafen.github.io/rbokeh/index.html#preview
library(rbokeh)
# prepare data
elements <- subset(elements, !is.na(group))
elements$group <- as.character(elements$group)
elements$period <- as.character(elements$period)

# add colors for groups
metals <- c("alkali metal", "alkaline earth metal", "halogen",
  "metal", "metalloid", "noble gas", "nonmetal", "transition metal")
colors <- c("#a6cee3", "#1f78b4", "#fdbf6f", "#b2df8a", "#33a02c",
  "#bbbb88", "#baa2a6", "#e08e79")
elements$color <- colors[match(elements$metal, metals)]
elements$type <- elements$metal

# make coordinates for labels
elements$symx <- paste(elements$group, ":0.1", sep = "")
elements$numbery <- paste(elements$period, ":0.8", sep = "")
elements$massy <- paste(elements$period, ":0.15", sep = "")
elements$namey <- paste(elements$period, ":0.3", sep = "")

# create figure
p <- figure(title = "Periodic Table", tools = c("resize", "hover"),
  ylim = as.character(c(7:1)), xlim = as.character(1:18),
  xgrid = FALSE, ygrid = FALSE, xlab = "", ylab = "",
  height = 445, width = 800) %>%

# plot rectangles
ly_crect(group, period, data = elements, 0.9, 0.9,
  fill_color = color, line_color = color, fill_alpha = 0.6,
  hover = list(name, atomic.number, type, atomic.mass,
    electronic.configuration)) %>%

# add symbol text
ly_text(symx, period, text = symbol, data = elements,
  font_style = "bold", font_size = "10pt",
  align = "left", baseline = "middle") %>%

# add atomic number text
ly_text(symx, numbery, text = atomic.number, data = elements,
  font_size = "6pt", align = "left", baseline = "middle") %>%

# add name text
ly_text(symx, namey, text = name, data = elements,
  font_size = "4pt", align = "left", baseline = "middle") %>%

# add atomic mass text
ly_text(symx, massy, text = atomic.mass, data = elements,
  font_size = "4pt", align = "left", baseline = "middle")

p
```


```{r htmlwidgets_chartype_table}
htmlwidgets_charttype_df <- data.frame(
  "Library" = c("Leaflet","dygraphs","Plotly", "rbokeh"),
  "URL" = c("http://rstudio.github.io/leaflet/",
            "http://rstudio.github.io/dygraphs",
            "https://github.com/ropensci/plotly",
            "http://hafen.github.io/rbokeh"),
  "Charts" = c(0, 0, 1, 1),
  "Scattergeo" = c(1, 0, 1, 1),
  "Choropleth" = c(1, 0, 1, 1),
  "Network" = c(0, 0, 1, 1),
  "Diagrams" = c(0, 0, 0, 1),
  "Time Series" = c(0, 0, 1, 1)
)

htmlwidgets_charttype_df$Library.URL <- paste0("<a href=",htmlwidgets_charttype_df$URL,">",htmlwidgets_charttype_df$Library,"</a>")

htmlwidgets_charttype_df[htmlwidgets_charttype_df==1] <- "<span class='glyphicon glyphicon-ok' aria-hidden='true'></span>"
htmlwidgets_charttype_df[htmlwidgets_charttype_df==0] <- "<span class='glyphicon glyphicon-remove' aria-hidden='true'></span>"

# Commented useful code
# colnames(htmlwidgets_charttype_df)[!grepl("^Library$|^URL$", colnames(htmlwidgets_charttype_df))]

htmlwidgets_charttype_df <- htmlwidgets_charttype_df[,c("Library.URL",colnames(htmlwidgets_charttype_df)[!grepl("Library|URL", colnames(htmlwidgets_charttype_df))])]

library(knitr)
kable(htmlwidgets_charttype_df, col.names = sub(pattern = "[.]", replacement = " ", colnames(htmlwidgets_charttype_df)))
```

### Explicit chart types

```{r explicit_chart_types}

charttypes_categorical_data <- c("Piechart","Barchart")

chart_types <- list(
  "Chart Type" = c("Piechart","Barchart","Violin Chart", "Box Whisker Chart", 
                 "Beeswarm Chart", "Gantt Chart", "Scatter Geo", "Chorpleth",
                 "Scatter Plot","Line Plot", "Function Plot", "Filled Area Chart",
                 "Bubble Charts", "Error Bars"),
  "Categorical Data" = c(),
  "Distributional Comparison" = c(),
  "Statistical" = c(),
  "Network" = c()
)
```

## Support for Customisability

Can this section include notes on creating annotations etc, a la https://plot.ly/r/shapes/

## Types of interactivity

```{r htmlwidgets_interactivitytype_table}
htmlwidgets_interactivitytype_df <- data.frame(
  "Library" = c("Leaflet","dygraphs","Plotly", "rbokeh"),
  "URL" = c("http://rstudio.github.io/leaflet/",
            "http://rstudio.github.io/dygraphs",
            "https://github.com/ropensci/plotly",
            "http://hafen.github.io/rbokeh"),
  "Scroll Zoom" = c(1, 1, 1, 1),
  "Area Select Zoom" = c(0, 1, 1, 1),
  "Panning" = c(1, 1, 1, 1),
  "Hover" = c(1, 1, 1, 1),
  "Shiny Click Events" = c(1, "unknown", "devel branch", "unknown")
)

htmlwidgets_interactivitytype_df$Library.URL <- paste0("<a href=",htmlwidgets_interactivitytype_df$URL,">",htmlwidgets_interactivitytype_df$Library,"</a>")

htmlwidgets_interactivitytype_df[htmlwidgets_interactivitytype_df==1] <- "<span class='glyphicon glyphicon-ok' aria-hidden='true'></span>"
htmlwidgets_interactivitytype_df[htmlwidgets_interactivitytype_df==0] <- "<span class='glyphicon glyphicon-remove' aria-hidden='true'></span>"

htmlwidgets_interactivitytype_df <- htmlwidgets_interactivitytype_df[,c("Library.URL",colnames(htmlwidgets_interactivitytype_df)[!grepl("Library|URL", colnames(htmlwidgets_interactivitytype_df))])]

library(knitr)
kable(htmlwidgets_interactivitytype_df, col.names = sub(pattern = "[.]", replacement = " ", colnames(htmlwidgets_interactivitytype_df)))
```

# Map Comparisons

What kind of maps?

# rbokeh

## Features

- Rectangle Zoom

## Geomaps

```{r rbokeh_scattergeo}
library(maps)
library(rbokeh)
data(world.cities)
caps <- subset(world.cities, capital == 1)
caps$population <- prettyNum(caps$pop, big.mark = ",")
figure(width = 800, padding_factor = 0) %>%
  ly_map("world", col = "gray") %>%
  ly_points(long, lat, data = caps, size = 5,
    hover = c(name, country.etc, population))
```



